%---------------------------------------------------------------------------%
%->> 封面信息及生成
%---------------------------------------------------------------------------%
%-
%-> 中文封面信息
%-
\confidential{}% 密级：只有涉密论文才填写
\schoollogo{scale=0.2}{ShanghaiTechLogo}% 校徽
%======论文题目 页眉显示 务必准确========
\title{面向动态目标的机械臂模仿学习控制方法研究}% 论文中文题目
\author{唐林峥}% 论文作者
\ID{2022533087}
\entranceYear{2022}
\advisor{白卫邦、马成毓}% 指导教师：姓名
\advisorsec{}% 第二指导老师：按情况填写
\degree{学士}% 学位：学士、硕士、博士
\degreetype{工学}% 学位类别：理学、工学、工程、医学等
\major{计算机科学与技术}% 二级学科专业名称
\institute{信息科学与技术学院}% 院系名称
\chinesedate{2026~年~6~月}% 毕业日期：夏季为6月、冬季为12月
%-
%-> 英文封面信息
%-
\englishtitle{Imitation Learning-Based Control of Robotic Manipulators for Dynamic Targets}% 论文英文题目
\englishauthor{Tang Linzheng}% 论文作者
\englishadvisor{Weibang Bai}% 指导教师
\englishdegree{Bachelor}% 学位：Bachelor, Master, Doctor。封面格式将根据英文学位名称自动切换，请确保拼写准确无误
\englishdegreetype{Engineering}% 学位类别：Philosophy, Natural Science, Engineering, Economics, Agriculture 等
\englishthesistype{thesis}% 论文类型： thesis, dissertation
\englishmajor{Computer Science and Technology}% 二级学科专业名称
\englishinstitute{School of Information Science and Technology}% 院系名称
\englishdate{June, 2026}% 毕业日期：夏季为June、冬季为December
%-
%-> 生成封面
%-
\maketitle% 生成中文封面
\makeenglishtitle% 生成英文封面
%-
%-> 作者声明
%-
\makedeclaration% 生成声明页
%-
%-> 中文摘要
%-
\chapter*{摘\quad 要}\chaptermark{摘\quad 要}% 摘要标题
\setcounter{page}{1}% 开始页码
\pagenumbering{Roman}% 页码符号

动态目标抓取要求机械臂在目标位姿和速度持续变化的条件下完成接近、相对对齐、闭爪接触和目标运输。模型预测控制（model predictive control, MPC）等基于模型的方法在状态观测准确时具有较强可解释性和约束处理能力，但在含噪感知输入下容易将位姿抖动传递到参考轨迹和闭爪判定中。模仿学习能够从专家示范中学习动态拦截和阶段决策规律，但其闭环性能取决于数据表示、动作输出形式以及对观测退化的鲁棒性。针对这一问题，本文在Genesis仿真平台中研究动态立方体抓取的目标中心化模仿学习方法，并在统一执行器下比较确定性动作块策略和生成式动作策略的闭环表现。

本文首先定义包含平面平移和定轴自旋的动态抓取基准任务，并以有限状态机结合MPC的FSM-MPC算法专家生成合成示教数据。随后，本文将抓取过程表示为目标中心化的历史相对观测、短时未来目标条件和相对任务空间动作块，使学习策略聚焦于相对位姿对齐和闭爪时机，而由共享MPC执行器处理底层约束跟踪。基于该统一接口，本文评估行为克隆（BC）、循环行为克隆（BC-RNN）、动作分块Transformer（ACT）和条件扩散策略四类模仿学习方法，并在真值观测、分解运动和含噪观测条件下与FSM-MPC专家进行比较。实验结果显示，在真值完整场景的750回合评估中，四类学习策略的成功率集中在86.0\%--87.2\%的狭窄区间内，均能形成有效闭环控制；FSM-MPC专家达到97.1\%，体现了显式阶段逻辑在无噪声条件下的显著上限优势。含噪评估进一步表明，四类模仿策略在高斯扰动和强组合噪声下均明显优于直接依赖原始噪声观测的未滤波FSM-MPC专家；但加入简单Kalman滤波后，FSM-MPC专家在Gaussian-only moderate和Full moderate噪声下分别恢复到94.0\%和85.0\%成功率。上述结果表明，目标中心化表示、历史条件和动作块模仿可以在未显式建模状态估计时提供一定的时间平滑作用，但其优势依赖噪声类型和基线配置；对模型控制基线而言，显式状态估计仍然至关重要。

\keywords{模仿学习，行为克隆，动作分块，动态目标抓取，模型预测控制}% 中文关键词
%-
%-> 英文摘要
%-
\chapter*{Abstract}\chaptermark{Abstract}% 摘要标题

Dynamic target grasping requires a manipulator to approach, align with, close on, and transport an object while the target pose and velocity continuously change. Model predictive control (MPC) provides interpretable constraint-aware execution under accurate state feedback, but noisy perception can be directly propagated into reference motion and grasp timing. Imitation learning offers a complementary route by learning interception and phase-selection behavior from expert demonstrations, yet its closed-loop performance depends strongly on state representation, action parameterization, and robustness to observation degradation. This thesis studies target-centric imitation learning for dynamic cube grasping in the Genesis simulation platform and compares deterministic and generative action-chunk policies under a shared MPC executor.

The proposed formulation represents each grasping segment using target-centric relative observation history, short-horizon target-motion conditions, and relative task-space action chunks. An FSM-MPC algorithmic expert provides synthetic demonstrations, while the learned policy predicts grasp-stage references and the shared MPC executor handles low-level constrained tracking. Within this unified interface, behavior cloning (BC), recurrent behavior cloning (BC-RNN), Action Chunking Transformer (ACT), and a conditional diffusion policy are evaluated under ground-truth, decomposed-motion, and noisy-observation settings. On the full ground-truth scene evaluated over 750 episodes, all four learned policies achieve success rates within a narrow 86.0\%--87.2\% band, while the FSM-MPC expert reaches 97.1\%, demonstrating a clear advantage of explicit phase logic when observations are accurate. Under noisy observations, learned action-chunk policies substantially outperform the unfiltered FSM-MPC expert under Gaussian jitter and strong combined noise; however, adding a simple Kalman filter recovers the FSM-MPC expert to 94.0\% under Gaussian-only moderate noise and 85.0\% under full moderate noise. These results suggest that target-centric action-chunk imitation can provide temporal smoothing when no explicit estimator is used, but its advantage depends on the noise type and baseline configuration; explicit state estimation remains essential for a strong model-based baseline.

\englishkeywords{Imitation Learning, Behavior Cloning, Action Chunking, Dynamic Grasping, MPC}% 英文关键词
%---------------------------------------------------------------------------%
